---
title: Exploration of 2016 US Presidential Campaign Contribution Data for the State of New York
author: "Roberto Zanchi"
date: "April 16, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<br>
This project explores a dataset of 2016 US presidential campaign contributions 
from donors in the State of New York. The data is sourced from
<http://fec.gov/disclosurep/pnational.do>.

#### 1. Load data and R packages
I start by importing the csv data downloaded from fec.gov into a data frame and
loading the R packages required for analysis and visualization.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
contribution_data <- read.csv("NY.csv")
library(dplyr)
library(ggplot2)
library(magrittr)
```

#### 2. Explore dataset structure
The dataset includes 121,529 observations of 19 variables, as shown by running
`str()` below. The only numerical variables are contribution amount
`contb_receipt_amt` and file number `file_num`, the other 17 variables being
categorical.

```{r str}
str(contribution_data)
```

#### 3. Explore variables of interest
In this section I explore a selection of variables that have caught my attention.

The number of observations - 121,529 - corresponds to the number of individual
financial contributions made.

I run `summary()` to obtain count values for categorical variables, and quartile
and mean statistics for numerical variables.

```{r summary, echo=TRUE}
summary(contribution_data)
```

I decide to further explore the following variables:

* candidate's name (`cand_nm`)
* contributor's city (`contbr_city`)
* contributor's occupation (`contbr_occupation`)
* contributor's employer (`contbr_employer`)
* contribution amount (`contb_receipt_amt`)
* contribution date (`contb_receipt_dt`)

##### Candidates
**Who are the candidates who have received campaign contributions from within \n
the State of New York?**

```{r candidates, echo=TRUE}
levels(contribution_data$cand_nm)
```

These 22 presidential candidates have received campaign contributions in the
State of New York in this dataset.

The list includes all candidates who are running active campaigns as of April 
2016: Clinton and Sanders for the Democratic Party, and Cruz, Kasich and Trump
for the Republican Party. The dataset does not include information about which
party the candidates belong to.

**How many individual campaign contributions have been made to each of the \n
candidates?**

```{r count, echo=FALSE}
ggplot(data = contribution_data, aes(x = contribution_data$cand_nm)) +
        geom_bar() +
        theme(text = element_text(size=12), 
              axis.text.x = element_text(angle=90, vjust=.5)) +
        xlab('Presidential candidates') +
        ylab('Number of contributions received')
```

Clinton and Sanders have received the greatest number of individual
contributions. Sanders has over 50,000 contributions and Clinton almost 45,000.
Carson and Cruz have received between 5,000 and 10,000. All other candidates
have received less than 5,000 each. 

##### Cities
**Which cities in the State of New York made the greatest number of \n
contributions?**

```{r}
head(summary(contribution_data$contbr_city))
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Top 10 contributors' cities by number of contributions
top_cities = contribution_data %>%
                group_by(contbr_city) %>%
                tally() %>%
                arrange(desc(n)) %>%
                top_n(10)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(top_cities, aes(x=contbr_city, y=n)) +
        geom_bar(stat='identity') +
        theme(text = element_text(size=12), 
              axis.text.x = element_text(angle=90, vjust=.5)) +
        xlab('Contributor city') +
        ylab('Number of contributions received')
```

New York City leads in number of individual contributions with 38,229. A quick
look at the data suggests that the actual value is likely to be larger, as
Brooklyn, The Bronx and Staten Island - reported separately in the dataset - are
New York City boroughs.

##### Occupations
**What are the occupations who have made the greatest number of contributions?**

The supporters in the State of New York who have made the greatest number of
campaign contributions are those retired, with over 15,000 contributions and not
employed, with over 12,500 contributions.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Top 10 occupations by number of contributions ex. retired and not employed
all_occupations = contribution_data %>%
                filter(contbr_occupation != '',
                       contbr_occupation != 'INFORMATION REQUESTED',
                       contbr_occupation !=
                               'INFORMATION REQUESTED PER BEST EFFORTS') %>%
                group_by(contbr_occupation) %>%
                tally() %>%
                arrange(desc(n)) %>%
                top_n(10)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=contbr_occupation, y=n), data=all_occupations) +
        geom_bar(stat='identity') +
        theme(text = element_text(size=12), 
              axis.text.x = element_text(angle=90, vjust=.5)) +
        xlab('Occupation of contributor') +
        ylab('Number of contributions')
```

I replotted the top 10 occupations after excluding individuals who are retired
and not employed. In the new top 10, supporters who are in legal professions
(attorneys and lawyers) and teachers or professors have made the largest number
of contributions.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Top 10 occupations by no. of contributions excluding retired and not employed
employed_occupations = contribution_data %>%
        filter(contbr_occupation != '',
               contbr_occupation != 'INFORMATION REQUESTED',
               contbr_occupation != 'INFORMATION REQUESTED PER BEST EFFORTS',
               contbr_occupation != 'RETIRED',
               contbr_occupation != 'NOT EMPLOYED') %>%
        group_by(contbr_occupation) %>%
        tally() %>%
        arrange(desc(n)) %>%
        top_n(10)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=contbr_occupation, y=n), data=employed_occupations) +
        geom_bar(stat='identity') +
        theme(text = element_text(size=12),
              axis.text.x = element_text(angle=90, vjust=.5)) +
        xlab('Occupation of contributor') +
        ylab('Number of contributions')
```

**Who are the top 10 employers by number of contributions of their employees?**

The top 10 employers of contributors are Clinton's campaign "Hillary for
America", universities (City University, Cornell, Columbia, NYU), the public
sector (City of New York, New York State, NYC Department of Education), Google
and Goldman Sachs.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Top 10 occupations by number of contributions ex. retired and not employed
top_employers = contribution_data %>%
                filter(contbr_employer != '',
                       contbr_employer != 'FREELANCE',
                       contbr_employer != 'HOMEMAKER',
                       contbr_employer != 'RETIRED',
                       contbr_employer != 'SELF-EMPLOYED',
                       contbr_employer != 'NOT EMPLOYED',
                       contbr_employer != 'N/A',
                       contbr_employer != 'NONE',
                       contbr_employer != 'SELF',
                       contbr_employer != 'SELF EMPLOYED',
                       contbr_employer != 'INFORMATION REQUESTED',
                       contbr_employer !=
                               'INFORMATION REQUESTED PER BEST EFFORTS') %>%
                group_by(contbr_employer) %>%
                tally() %>%
                arrange(desc(n)) %>%
                top_n(10)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=contbr_employer, y=n), data=top_employers) +
        geom_bar(stat='identity') +
        theme(text = element_text(size=12),
              axis.text.x = element_text(angle=90, vjust=.5)) +
        xlab('Employer of contributor') +
        ylab('Number of contributions')
```

##### Contribution amount
**How can all the available contribution amount data be described?**

Summary statistics of the complete set of 121,529 observations show:

* Prevalence of contributions of small value (median of USD 50 and USD 116 as
3rd Quartile);
* Mean contribution amount (US$365) higher than 3rd Quartile, suggesting long
tail of values;
* A maximum value of USD 3,686,000, confirming the existence of outliers;
* Presence of negative values (probably refunds), e.g. the minimum of USD -5,400.

```{r}
summary(contribution_data$contb_receipt_amt)
```

I visualize this data through a density plot, applying a log10 transformation to
positive `contb_receipt_amt` values on the x axis and a sqrt transformation to
density on the y-axis. The plot resembles a normal distribution with a long tail
including values around 1e+06 (i.e. >USD 1m).

```{r, echo=FALSE, message=FALSE, warning=FALSE}
positive_contributions = subset(contribution_data,
                                contribution_data$contb_receipt_amt > 0)
# Plot density of contribution amounts
ggplot(positive_contributions, aes(x=contb_receipt_amt)) +
        geom_density() +
        scale_x_log10() +
        scale_y_sqrt() +
        xlab('Contribution amount (USD) (log10)') +
        ylab('Density (sqrt)')
```

##### Contribution dates
**What is the number of contributions over time?**

In order to analyse and visualize contribution dates, I convert date values in
the dataset from string (e.g. "29-FEB-16") to Date format (e.g. 2016-02-29) and
select observations with positive contribution amounts and non-NA date values.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Convert dates from strings to the Date datatype
contribution_data$contb_receipt_dt = as.Date(contribution_data$contb_receipt_dt,
                                             "%d-%b-%y")
# Omitt negative amounts
selected_dates = filter(contribution_data, contb_receipt_amt > 0,
                        !is.na(contb_receipt_dt))
```

The filtered dataset contains 120,068 observations with contribution dates
ranging from 11 October 2013 and 29 February 2016.

```{r}
summary(selected_dates$contb_receipt_dt)
```

The frequency plot below shows how the number of individual contributions has
grown over time. Exponential growth in number of contributions started around
April 2015. Quartile statistics and the plot show that 75% of all contributions
were received starting October 2015 and more than half starting January 2016.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Time series with count of contributions from October 2013 to February 2016
ggplot(selected_dates, aes(x=contb_receipt_dt))+ 
        geom_freqpoly() +
        xlab('Date of contribution') +
        ylab('Count of individual contributions')
```

#### 4. Explore relationships between variables
In this section I explore relationships between different variables.

##### Data wrangling
Before doing further analysis, I select the observations that meet the following
criteria:

* Non-negative contribution amount;
* Max contribution amount of USD 2,700;
* Non-NA contribution date value.

According to [US Federal Election Commission rules](http://www.fec.gov/info/contriblimitschart1516.pdf)
, USD 2,700 is the contribution limit for individuals' donations to candidates. 

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Select based on criteria
select_data = filter(contribution_data,
                     contb_receipt_amt > 0 & contb_receipt_amt <= 2700 &
                     !is.na(contb_receipt_dt))
# Convert dates from strings to the Date datatype
select_data$contb_receipt_dt = as.Date(select_data$contb_receipt_dt, "%d-%b-%y")
```

##### Contributions in the filtered dataset
**How can the contribution amount data in the filtered dataset be described?**

The new data frame `select_data` has 119,785 observations.

```{r}
length(select_data$contb_receipt_amt)
```

The total value of reported contributions in the period October 2013 - February
2016 is USD 35,911,638.

```{r}
sum(select_data$contb_receipt_amt)
```

The majority of contributions is in small amounts. The median is USD 50 and the
3rd Quartile USD 121.90 over the entire period considered.

```{r}
summary(select_data$contb_receipt_amt)
```

As the plot shows, in the 4th Quartile there are certain common donation
amounts: USD 250, 500, 1,000 and 2,700 (the maximum allowed).

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(select_data, aes(x=contb_receipt_amt)) +
        geom_density() +
        xlab('Contribution amount (USD)') +
        ylab('Density')
```

##### Value of contributions over time
**What is the value of contributions over time?** 

Most of the donations from within the State of New York were made after April
2015, which is the month both Clinton and Sanders formally [announced their candidacy](https://en.wikipedia.org/wiki/United_States_presidential_election,_2016_timeline#April).

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(select_data, aes(x=contb_receipt_dt, y=contb_receipt_amt)) +
        geom_point(alpha = 1/50) +
        xlab('Date of contribution') +
        ylab('Contribution amount (USD)')
```

To understand the actual monetary value of the donations over time, I plot a
graph with mean donation values by date.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(select_data, aes(x=contb_receipt_dt, y=contb_receipt_amt)) +
        stat_summary(fun.y="mean", geom="bar") +
        xlab('Date of contribution') +
        ylab('Contribution amount (USD (mean)')
```

While the number of individual contributions has increased over time, the mean
value of contributions has consistently decreased over time, from mean values of
nearly USD 2,000 in April 2014 to less than USD 200 in February 2016. Most of
the highest value contributions were made right after the time of candidacy
announcement for Clinton and Sanders in April 2015.

##### Value of contributions by city
**What is the total value of contributions for the top 10 cities?**

I explore the value of contributions by city by plotting the total value
(sum of `contb_receipt_amt`) of individual contributions for the top 10 cities
by number of contributions.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Get contributions by occupation
top_cities_value = filter(select_data,
                          contbr_city %in%
                                  top_cities[['contbr_city']])
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(top_cities_value, aes(x=contbr_city, y=contb_receipt_amt)) +
        stat_summary(fun.y="sum", geom="bar") +
        xlab('Cities') +
        ylab('Total contributions (USD)') +
        theme(legend.position="none", 
              axis.text.x = element_text(angle=90, vjust=.5))
```

New York City contributed almost USD 20 million of the total USD 35,911,63 in
the filtered dataset. Another USD 2.5 million were contributes by supporters in
Brooklyn, also part of New York City.

##### Number of contributions by candidate
For this exploration, I select the candidates who are still running active
campaigns: Clinton, Cruz, Kasich, Sanders and Trump.

```{r}
selected_candidates = filter(select_data,
                        cand_nm == "Clinton, Hillary Rodham" |
                                cand_nm == "Cruz, Rafael Edward 'Ted'" |
                                cand_nm == "Kasich, John R." |
                                cand_nm == "Sanders, Bernard" |
                                cand_nm == "Trump, Donald J.")
```

**How many contributions has each candidate received in the period?**

This bar chart is a subset of the first graph

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(selected_candidates, aes(x=cand_nm)) +
        geom_bar(aes(fill = cand_nm)) +
        theme(legend.position="none") +
        theme(text = element_text(size=12), 
              axis.text.x = element_text(angle=90, vjust=.5)) +
        xlab('Presidential candidates') +
        ylab('Number of contributions received')
```

**How is number of contributions distributed over the period for each candidate?**

I use a density diagram to see how each candidate's total contributions are
distributed over the period. 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Time series density plot with contributions from October 2013 to February 2016
ggplot(selected_candidates, aes(x=contb_receipt_dt)) +
        geom_density(aes(color = cand_nm)) +
        xlab('Date of contribution') +
        ylab('Density') +
        theme(legend.title=element_blank(), 
              legend.text = element_text(size = 8),
              legend.position="bottom",
              legend.direction="horizontal")
```

##### Value of contributions by candidate
**What is the value of the contributions received by each candidate?**

The mean value of contributions received by candidates in the overall period is
plotted below. Kasich, who had one of the smallest numbers of contributions, has
the highest mean value of almost USD 1,000. He is followed by Clinton with an
average contribution value of almost USD 500. Sanders has the smallest mean
value of contributions, less than USD 100.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(selected_candidates, aes(x=cand_nm, y=contb_receipt_amt)) +
        stat_summary(fun.y="mean", geom="bar", 
                     aes(fill = selected_candidates$cand_nm)) +
        xlab('Candidates') +
        ylab('Contribution amount (USD) (mean)') +
        theme(legend.position="none", 
              axis.text.x = element_text(angle=90, vjust=.5))
```

I then use boxplots to explore the distribution of the value of contribution
received by each candidate.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=cand_nm, y=contb_receipt_amt), data=selected_candidates) +
        geom_boxplot(aes(fill = cand_nm)) +
        xlab('Candidates') +
        ylab('Contribution amount (USD) (boxplot)') +
        theme(legend.position="none") +
        theme(text = element_text(size=10), 
              axis.text.x = element_text(angle=90, vjust=.5))
```

The boxplots show that most of the contributions have small median value and are
concentrated around the median for Clinton, Cruz, Sanders and Trump. While
having received few contributions, Kasich has the highest value median
contribution.

##### Contributions by occupation
**What is the value of the contributions made by occupation?**

In order to explore the value of contributions by occupation I plot their mean
values.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Get contributions by occupation
contbs_top_occups = filter(select_data,
                           contbr_occupation %in%
                               all_occupations[['contbr_occupation']])
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(contbs_top_occups, aes(x=contbr_occupation, y=contb_receipt_amt)) +
        stat_summary(fun.y="mean", geom="bar") +
        xlab('Occupations') +
        ylab('Contribution amount (USD (mean)') +
        theme(legend.position="none", 
              axis.text.x = element_text(angle=90, vjust=.5))
```

Homemakers have the highest mean contribution value of USD 1,000, followed by
lawyers and attorneys each with mean contributions of over USD 500, and
consultants with mean contribution just below USD 500.

Boxplots then allow us to explore the distribution of contribution value by
occupation.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Box plot contributions by occupation
ggplot(contbs_top_occups, aes(x=contbr_occupation, y=contb_receipt_amt)) +
    geom_boxplot() +
    theme(axis.text.x = element_text(angle = 90, vjust = .5)) +
        ylab('Contribution amount (USD) (boxplot)') +
        xlab('Occupation of contributor')
```

Housemakers also have the highest median contribution.

#### 5. Final plots and summary

In this section I present three of the plots explored in the previous section
with some additional comments.

##### 1. Trend in value of contributions
**As the number of contributions grows, their mean value is declining**

One of the research questions in the exploration was about how value of
contributions changed over time. To answer it, I  plotted a bar chart with mean
contributon amount for each day of the campaign, and a smoothed conditional mean
line to highlight the trend.

I discovered that, while the number of contributions per day has grown since
April 2015, their mean value has been steadily declining. A possible explanation
is that contributions of smaller amount have grown faster than the others.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(select_data, aes(x=contb_receipt_dt, y=contb_receipt_amt)) +
        stat_smooth() +
        stat_summary(fun.y="mean", geom="bar") +
        xlab('Date of contribution') +
        ylab('Contribution amount (USD) (mean)') +
        ggtitle("Mean contribution amount over time")
```

##### 2. Popularity of candidates
**All candidates have higher density of contributions between December 2015 \n 
and  February 2016**

Another key question was about how the number of contributions has varied over
time for each candidate. I used density plots to visualize how the number of
contributions is distributed over time, independently of their absolute value.

For all five selected candidates the density of contributions has risen since
December 2015. Sanders has a higher proporion of contributions made to him
between January and February 2016 than any other candidate.

Could this mean that the popularity of Sanders has grown faster than other
candidates' in the period? Further investigation would be required to establish
how this phenomenon is significantly related to popularity.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(selected_candidates, aes(x=contb_receipt_dt)) +
        geom_density(aes(color = cand_nm)) +
        xlab('Date of contribution') +
        ylab('Density of contributions for each candidate') +
        theme(legend.title=element_blank(),
              legend.text = element_text(size = 8),
              legend.position="bottom",
              legend.direction="horizontal") +
        ggtitle("Density plot of contributions for each candidate")
```

##### 3. The greatest contributors
**Housemakers have the highest mean and median contribution values**

A third question in the exploration was about how the value of contributions
made is related to the occupation of supporters. I visualized this through
boxplots and mean contribution value for the top 10 occupations by number of
contributions. Mean in blue and outliers in red.

Housemaker supporters have the highest median and mean contribution value of all
occupations in the top 10. All other professions also have smaller variance,
and mean values beyond the 3rd Quartile, as a result of outlier contribution
values up to the maximum of USD 2,700.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(contbs_top_occups, aes(x=contbr_occupation, y=contb_receipt_amt)) +
    geom_boxplot(outlier.colour = "red", outlier.size = .5) +
        stat_summary(fun.y=mean, geom="point", colour = "blue", size = 2) +
        theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) +
        ylab('Contribution distribution (boxplot) / mean (blue)') +
        xlab('Occupation of contributor') +
        ggtitle("Distribution and mean of contributions by occupation")
```

#### 6. Reflection

**Summary of the exploration process**

I decided to explore a dataset of financial contribution to 2016 US presidential
candidates made available by the US Federal Election Commission. I selected the
available data for the State of New York.

After loading an exploring the structure of the dataset in RStudio, I performed
univariate exploration of selected variables, followed by exploration between
selected variables.

In particular, I focused on:

* Number and value of the contributions;
* How number and value vary over time;
* The five candidates with active campaigns as of April 2016;
* Top 10 contributor occupations by number of contributions made;
* Top 10 cities by number of contributions made.

I then selected three plots and provided additional comments.

**Important choices made**

The most important choices I made with in data wrangling and analysis are:

* I excluded the contributions of value greater than USD 2,700;
* I focused my analysis on the five candidates with active campaigns;
* I excluded retired and non-employed contribitors from occupation statistics.

**Challenges**

The single greatest challenge in this data exploration has been working with a
large number of categorical variables, and only contribution value as a
numerical variable.

**Ideas for future research**

Ideas for future research could be:

* Extending the analysis to include multiple US states;
* Exploring the data about candidates who no longer have active campaigns, for
comparison.